---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# predictSource

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/benmarwick/predictSource.svg?branch=master)](https://travis-ci.org/benmarwick/predictSource)
<!-- badges: end -->

The package predictSource provides functions to verify that data can be used to separate sources of samples and then to predict the sources of additional samples.  Data can be both quantitative and qualitative.  The data used as examples are 5 elements from 113 obsidian samples from five sources in the Jemez caldera in north central New Mexico USA and the same 5 elements from 91 artifacts obtained from nearby archaeological sites.

The package contains functions for exploratory data analysis (descriptive statistics, 2- and 3-dimensional plots [the latter can be rotated], tests for 1- and 2-dimensional Gaussian distributions [helpful in identifying outliers]) and multivariate analysis (principal components, classification trees, and random forests).  A proposed analysis strategy is to use random forests to evaluate whether the data can separate the sources and to identify the most important predictors if there are many, use a classification tree to understand how the data are used to sources, use random forests to predict the sources of unknown samples, and then evaluate the validity of the predictions by ploting the first two principal components of the unknowns with the convex hulls of the known sources.  The random forest analysis also produces the probabilities of assignment to each source for each sample; this can be helpful in identifying unknowns that are difficult to classify.

Obsidian sources are usually predicted using 2- and 3-dimensional scatter plots.  This methodology is much faster and can identify errors in assignment (the example data includes one such error).  Archaological information should be used with any analysis procedure.

A detailed vignette provides examples for the use of each function (using the obsidian data) and some background for classification trees, random forests, and checking for Gaussian distributions.

## Run in Binder

[![Binder](http://mybinder.org/badge.svg)](http://mybinder.org/v2/gh/benmarwick/predictSource/master)

Click on the button above to launch RStudio in your browser so you can test or use this package without installing it locally on your computer. See these [instructions](https://github.com/rocker-org/binder#opening-rstudio-once-binder-launches) for full details on how to to open RStudio.

## Installation

You can install predictSource from GitHub with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("benmarwick/predictSource")
```

