---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# predictSource

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/benmarwick/predictSource.svg?branch=master)](https://travis-ci.org/benmarwick/predictSource)
<!-- badges: end -->

The package predictSource provides functions to verify that data can be used to separate sources of samples, to predict the sources of additional samples, and to create plots that evaluate the validity of the predictions.  Data can be both quantitative and qualitative.  A proposed analysis strategy is to use random forests to evaluate whether the data can separate the sources and to identify the most important predictors if there are many, use a classification tree to understand how the data are used to separate sources, use random forests to predict the sources of unknown samples, and then evaluate the validity of the predictions by ploting the first two principal components of the unknowns with the convex hulls of the known sources.  The random forest analysis also produces the probabilities of assignment to each source for each sample; this can be helpful in identifying unknowns that are difficult to classify.The package also contains functions for exploratory data analysis (descriptive statistics, 2- and 3-dimensional plots [the latter can be rotated], tests for 1- and 2-dimensional Gaussian distributions [helpful in identifying outliers]) and multivariate analysis (principal components). A detailed vignette provides examples for the use of each function (using obsidian data in the examples) and some background for classification trees, random forests, and checking for Gaussian distributions.

The motivation for the package was predicting the sources of obsidian artifacts. Archaeologists and geochemists usually do this using 2- and 3-dimensional scatterplots. The functions in this package make predictions much faster than can be done with scatterplots; the principal components graphic identifies objects that are or may be misclassified. Archaeological knowledge should also be used in making predictions.

## Run in Binder

[![Binder](http://mybinder.org/badge.svg)](http://mybinder.org/v2/gh/benmarwick/predictSource/master)

Click on the button above to launch RStudio in your browser so you can test or use this package without installing it locally on your computer. See these [instructions](https://github.com/rocker-org/binder#opening-rstudio-once-binder-launches) for full details on how to to open RStudio.

## Installation

You can install predictSource from GitHub with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("benmarwick/predictSource")
```

